Turbo fast no compromise gravitational wave parameter estimation

### 2022-12-21 src/tex/ms.tex
Grammarly

% Abstract with filler text
\begin{abstract}
We present a light-weighted, flexible and high performance framework to estimate
gravitational wave event parameters. By combining heterodyned likelihood,
automatically differentiable and accelerators compatible waveforms, normalizing
flow enhanced gradient-based MCMC, we achieve parameter estimation for real
event like GW150914 and GW170817 within a minute of sampling time with little to
no compromise. Our framework does not require pre-training or explicit
reparameterization, and can be generalized to handle higher dimensional
problems. We also discuss as near-real time parameter estimation have been shown
to be possible by multiple groups, what are the trade-offs and future
developments the community should be aware of. Our code for generating the
manuscript and running the analysis is publically available on github.
\end{abstract}

@@

% Brief description of PE and old generation of code
Parameter estimation (PE) is one of the most commonly perform data analysis
tasks in gravitational wave (GW) data analysis. The central question of PE is to
infer the parameters of the GW source given the strain data. The PE results in
the most recent catalog of GW events are powered by a number of community
developed PE codes, including \texttt{lalinference}, \texttt{pycbc}, and
\texttt{bilby} \cite{Romero-Shaw:2020owr} \kw{cite}. These packages have been tested by a number of groups
and well regarded as the standard tools.

% Emerging tools from different group
While the standard tools have passed many robustness tests, they are known to be
computationally intensive. The exact amount of time needed to analyze one event
depends on the length of the event. These community tools take about a week to
analyze shorter events such as binary black hole (BBH) mergers, where they can
take up to a month on a computer cluster to analyze longer events such as binary
neutron star (BNS) mergers. There are efforts from multiple groups developing
tools to speed up the PE process \kw{Cite}.

@@

In this work, we present a light-weighted, flexible and high performance
framework to estimate GW event parameters. Our framework implemneted the
following major features to achieve its performance:
\begin{enumerate}
\item Differential waveform models
\item Normalizing flow enhanced MCMC sampler
\item Heterodyned likelihood
\item Native support for accelerator
\item JIT compilation
\end{enumerate}
\kw{Dingo can't do BNS yet}

@@

The rest of the paper is structured as follows: We review the basic of PE and
introduce our framework in section \ref{sec: PE}. We present benchmarking
results on both simulated and real data in section \ref{sec: Result}. Finally,
we discuss the implications of this work and directions for future development
in section \ref{sec: Discussion}.

@@

The main objective in PE can be summarized as: given data in the form of a time
series of strain $\mathbf{d}$, find the distribution of the source parameters
$\mathbf{\theta}$ that is compatible with the data, i.e.
$p(\mathbf{\theta}|\mathbf{d})$. In order to compute this object, we can rewrite
it using Bayes' theorem,
\begin{align}
    p(\theta| d) = \frac{\mathcal{L}(d|\theta)\pi(\theta)}{p(d)},
\end{align}

where $\mathcal{L}(d|\theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distributions, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins.
Assuming the noise follows from a stationary Gaussian process, the log
likelihood in the case of GW is given by

\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right handside of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likleihood throughout this work, which can be defined as 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluting the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain using can takes a day to half a year, which makes NR prohibitably
expensive for PE. To circumvent this problem, there are a number of waveform
"approximant" families, including the IMRPhenom family \kw{}, SEOB family \kw{},
and NR surrogate family\kw{cite}. For shorter event such as a $30-30\ M_{\odot}$
BBH, one waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number
in lalsuite}. For longer event such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evalutation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in time or frequency domain, the
computational burden for longer-duration signals are often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\kw{cite}. We use heterodyned likelihood \kw{cite} (also named relative binning
in \kw{cite}) in this work. The idea behind heterodyned likelihood can be
summarized as the following: the integrand in eq. \ref{eq: innerproduct} is
highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that are
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that is depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

One may be concerned by the accuracy of this approximation over the target
parameter space, especially in the region where the generated waveform is
significantly different from the reference waveform. However, given we are
interested in the most probable set of parameters, if we choose the reference
waveform to be close to the data, the waveforms that are different from the
reference waveform should also differ significantly from the data. This means
the likelihood value for these waveforms should be significantly smaller than
the likelihood for the waveforms that are similar to the reference waveform,
hence will not be relevant for the PE result. In practice, one will first
optimize the likelihood function with full frequnecy resolution to obtain the
reference waveform parameters, which can be run at a much lower cost compared to
PE.

We give a concise description of what we implement
in our code here. For more extensive  derivation of heterodyned
likelihood, we refer the reader the reference \cite{Zackay:2018qdy}.
In the heterodyned likelihood framework, the two terms in eq \ref{eq:
loglikelihood} are given by
\begin{align}
    \langle d|h \rangle &\approx \sum_b A_0(b) r^*_0(h,b) + A_1(b) r^*_1(h,b), \nonumber \\
    \langle h|h \rangle &\approx \sum_b B_0(b) |r_0(h,b)|^2 + 2 B_1(b) \Re[r_0(h,b)r_1(h,b)],
\label{eq: heterodynedlikelihood}
\end{align}

@@

Considering the phasing of a waveform is denoted by a power series $\Psi(f) =
\sum_i \alpha_i f^{\gamma_i}$, where $\alpha_i$ are some coefficients depending
on the waveform parameters and $\gamma_i$ are powers motivated by post-Newtonian
thoery. For example, for the term $\gamma_i = -5/3$, $\alpha_i$ is related to
the chirp mass. The maximum dephasing one can have within a frequency interval $[f_{\textrm{min}},f_{\textrm{max}}]$ is given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i>=0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in eq. \ref{eq:
maxdephasing}, we can choose the binning scheme which devide the entire
frequency domain of interests into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

Given a sufficiently dense binning scheme, the ratio between the target waveform
and the reference within a bin can be approximated as by linear interpolation
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin resepctively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and invert eq. \ref{eq:definer}.

The final ingredient we need is the heterodyning coefficients given the data and
the reference waveform on the sparse bins, which are given by
\begin{align}
    A_0(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)}{S_n(f)} df, \\
    A_1(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)(f-f_m(b))}{S_n(f)} df, \\
    B_0(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2}{S_n(f)} df, \\
    B_1(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2(f-f_m(b))}{S_n(f)} df.
\end{align}
Note that the sum within each bins should be done with the same sampling rate as
the data, i.e. the same one would do in the case without using heterodyned
likelihood.

@@

In MCMC, the posterior distribution is approximated by a Markov chain that
converges to the target distribution. The Markov chain is constructed by
iteratively proposing a new point in the parameter space based on the current
location of the chain. The proposed point is accepted with a probability that is
usually set to be proportional to the ratio of the posterior density evaluated
at the proposed point and the current point. The chain can either accept the
proposal and move to the new location, or reject the proposal and stay at the
current location. This process is repeated until the chain converges to the
target distribution. The samples generated by the chain are then used to
estimate the quantities of interest, such as the mean and credible intervals of
the source parameters. In practice, since we do not know the target distribution
ahead of time, the MCMC process is usually repeated until a certain criterion is
met, such as a Gelman-Rubin convergence statistic \cite{Gelman-rhat} lower than
a threshold or simply after a certain number of iterations.

Compared to direct sampling, MCMC algorithms only explore regions that are
highly probable, thus reducing the computational cost by not wasting resource in
region where it is unlikely to generate the data. However, MCMC algorithms come
with their own set of issue. To illustrate what difficulties a MCMC algorithms
may face, we can examine one of the most vanilla MCMC algorithm, the
Metropolis-Hastings algorithm with a Gaussian kernel. Starting at some initial
point, one can draw a proposed point from a Gaussian transition kernel, defined
as
\begin{align}
    q(\mathbf{x},\mathbf{x_0})= \mathcal{N}(\mathbf{x}|\mathbf{x_0},\mathbf{C}),
\end{align}
where $\mathbf{x_0}$ is the current location of the chain,
$\mathbf{x}$ is the proposed location, and $\mathbf{C}$ is the
covariance matrix of the Gaussian. In the simplest case, we can pick
$\mathbf{C}$ to be a diagonal matrix with a constant value, which corresponds to
an isotropic Gaussian centers around the current location with a constant variance.
And the acceptance criteria is defined as
\begin{align}
\alpha(\mathbf{x},\mathbf{x_0}) = \min\left(1,\frac{p(\mathbf{x})q(\mathbf{x_0},\mathbf{x})}{p(\mathbf{x_0})q(\mathbf{x},\mathbf{x_0})}\right).
\label{eq:Gaussian_acceptance}
\end{align}

We can see from eq. \ref{eq:Gaussian_acceptance} the acceptance rate is
proportion to fraction of volume where the posterior density at the proposed
location is higher than the current location within the Gaussian transition
kernel. If we choose the variance of the transition kernel to be too large, this
fraction will be small hence the acceptance rate will be poor. On the other
hand, if one choose the variance to be too small, nearby samples will be
correlated. In both cases, the efficiency in constructing the chain with a
target number of independent samples is suboptimal. There is often a tuning
process before we run the MCMC algorithm to find the optimal tuning parameters,
in this example the variance of the Gaussian, to ensure close to optimal
performance.

However, as we are dealing with higher dimensional problems, even the optimally
tuned Gaussian transition kernel does not guarantee good performance. In order
to have a reasonable acceptance rate, the variance of the Gaussian has to be
smaller in a higher dimensional space, which means the transition kernel is in
general making smaller and smaller step as we increase the dimensionality.

Transition kernels which leverage gradient information of the target
distribution can help to address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluate at the current location to propose a new
point. For example, Metropolis-adjusted Langevin algorithm (MALA) \kw{cite}
place a unit Gaussian at the tip of the gradient vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I}),
\end{align}
where $\tau$ is the step size, which is a tuning parameter. Compared to a
Gaussian centers at the current location, the MALA transition kernel is more
likely to propose a point into the higher posterior density region because of
the gradient term, which helps to boost the acceptance rate.

While transition kernels that use gradient information can help to improve the
acceptance rate, computing the gradient of the posterior density function
introduce additional computational cost, which is not necessary beneficial in
terms of sampling time. If one wants to compute the gradient information through
finite differencing, the additional computation cost goes as at least $\sim
\mathcal{O}(2n)$, where $n$ is the dimension of the problem. \texttt{Jax} allows
us to compute the gradient of the likelihood function with respect to the
parameters through automatic differentiation, which gives the gradient
information down to machine precision at around the same order of time compared
to evaluating the posterior itself. Having access to gradient information
through automatic differentiation is crucial to make the trade-off between using
gradient-based transition kernels and additional computation cost favorable.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples\kw{cite}, there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first order derivative here. Sampling
algorithms that use information of higher order derivative such as manifold-MALA
and Riemannian-HMC can in principle decorrelate local correlation in the target
distribution, however they often have instability issue when they are used on
real-life application, so they are not used often in practice.} For examples,
target distribution that exhibits local correlation are hard to dealt with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely to
be trapped in one mode and take extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

% Long burn-in too
Moreover, before we can use the sampling chain to estimate the quantity we care,
the sampler often needs to first find the most probable region in the target
space, which is a common process that is referred to "burn-in" in the
literature. This means one would discard a certain amount of data generated from
the beginning and only use the later part to estimate the quantities of
interest. The burn-in phase of a gradient-based sampler is often as long as the
sampling phase, which means a good portion of computation is not necessary
helpful in estimating the target quantities.

% Crux of normalizing flow
Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution, and $f$ is an invertible parameterized transform that
connect the two distributions. For a detail discussion of the algorithm, we
refer the readers to \kw{cite}.

>>>>

% Abstract with filler text
\begin{abstract}
We present a light-weighted, flexible, and high-performance framework to estimate
gravitational wave event parameters. By combining heterodyned likelihood,
automatically differentiable and accelerators compatible waveforms, normalizing
flow enhanced gradient-based MCMC, we achieve parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time with little to
no compromise. Our framework does not require pre-training or explicit
reparameterization and can be generalized to handle higher dimensional
problems. We also discuss as near real-time parameter estimation has been shown
to be possible by multiple groups, what are the trade-offs and future
developments the community should be aware of. Our code for generating the
manuscript and running the analysis is publically available on GitHub.
\end{abstract}

@@

% Brief description of PE and old generation of code
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational wave (GW) data analysis. The central question of PE is to
infer the parameters of the GW source given the strain data. The PE results in
the most recent catalog of GW events are powered by several community-developed
PE codes, including \texttt{lalinference}, \texttt{pycbc}, and \texttt{bilby}
\cite{Romero-Shaw:2020owr} \kw{cite}. These packages have been tested by a
number of groups and are well regarded as the standard tools.

% Emerging tools from different group
While the standard tools have passed many robustness tests, they are known to be
computationally intensive. The exact amount of time needed to analyze one event
depends on the length of the event. These community tools take about a week to
analyze shorter events such as binary black hole (BBH) mergers, whereas they can
take up to a month on a computer cluster to analyze longer events such as binary
neutron star (BNS) mergers. There are efforts from multiple groups developing
tools to speed up the PE process \kw{Cite}.

@@

In this work, we present a light-weighted, flexible, and high performance
framework to estimate GW event parameters. Our framework implements the
following major features to achieve its performance:
\begin{enumerate}
\item Differential waveform models
\item Normalizing flow enhanced MCMC sampler
\item Heterodyned likelihood
\item Native support for accelerator
\item JIT compilation
\end{enumerate}
\kw{Dingo can't do BNS yet}

@@

The rest of the paper is structured as follows: We review the basics of PE and
introduce our framework in section \ref{sec: PE}. We present benchmarking
results on both simulated and real data in section \ref{sec: Result}. Finally,
we discuss the implications of this work and directions for future development
in section \ref{sec: Discussion}.

@@

The main objective in PE can be summarized as given data in the form of a time
series of strain $\mathbf{d}$, find the distribution of the source parameters
$\mathbf{\theta}$ that is compatible with the data, i.e.
$p(\mathbf{\theta}|\mathbf{d})$. To compute this object, we can rewrite
it using Bayes' theorem,
\begin{align}
    p(\theta| d) = \frac{\mathcal{L}(d|\theta)\pi(\theta)}{p(d)},
\end{align}

where $\mathcal{L}(d|\theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins.
Assuming the noise follows from a stationary Gaussian process, the
log-likelihood in the case of GW is given by

\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right hand side of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likelihood throughout this work, which can be defined as 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluating the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain can take a day to half a year, which makes NR prohibitively
expensive for PE. To circumvent this problem, there are several waveform
"approximant" families, including the IMRPhenom family \kw{}, SEOB family \kw{},
and NR surrogate family\kw{cite}. For shorter events such as a $30-30\ M_{\odot}$
BBH, one waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number
in lalsuite}. For longer events such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evaluation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in the time or frequency domain, the
computational burden for longer-duration signals is often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\kw{cite}. We use heterodyned likelihood \kw{cite} (also named relative binning
in \kw{cite}) in this work. The idea behind heterodyned likelihood can be
summarized as the following: the integrand in eq. \ref{eq: innerproduct} is
a highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that is
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

One may be concerned by the accuracy of this approximation over the target
parameter space, especially in the region where the generated waveform is
significantly different from the reference waveform. However, given we are
interested in the most probable set of parameters, if we choose the reference
waveform to be close to the data, the waveforms that are different from the
reference waveform should also differ significantly from the data. This means
the likelihood value for these waveforms should be significantly smaller than
the likelihood of the waveforms that are similar to the reference waveform,
hence will not be relevant for the PE result. In practice, one will first
optimize the likelihood function with full frequency resolution to obtain the
reference waveform parameters, which can be run at a much lower cost compared to
PE.

We give a concise description of what we implement
in our code here. For a more extensive derivation of heterodyned
likelihood, we refer the reader to the reference \cite{Zackay:2018qdy}.
In the heterodyned likelihood framework, the two terms in eq \ref{eq:
loglikelihood} are given by
\begin{align}
    \langle d|h \rangle &\approx \sum_b A_0(b) r^*_0(h,b) + A_1(b) r^*_1(h,b), \nonumber \\
    \langle h|h \rangle &\approx \sum_b B_0(b) |r_0(h,b)|^2 + 2 B_1(b) \Re[r_0(h,b)r_1(h,b)],
\label{eq: heterodynedlikelihood}
\end{align}

@@

Considering the phasing of a waveform is denoted by a power series $\Psi(f) =
\sum_i \alpha_i f^{\gamma_i}$, where $\alpha_i$ are some coefficients depending
on the waveform parameters and $\gamma_i$ are powers motivated by post-Newtonian
theory. For example, for the term $\gamma_i = -5/3$, $\alpha_i$ is related to
the chirp mass. The maximum dephasing one can have within a frequency interval $[f_{\textrm{min}},f_{\textrm{max}}]$ is given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i>=0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in eq. \ref{eq:
maxdephasing}, we can choose the binning scheme which divide the entire
frequency domain of interests into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

Given a sufficiently dense binning scheme, the ratio between the target waveform
and the reference within a bin can be approximated by linear interpolation
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin respectively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and invert eq. \ref{eq:definer}.

The final ingredient we need is the heterodyning coefficients given the data and
the reference waveform on the sparse bins, which are given by
\begin{align}
    A_0(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)}{S_n(f)} df, \\
    A_1(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)(f-f_m(b))}{S_n(f)} df, \\
    B_0(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2}{S_n(f)} df, \\
    B_1(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2(f-f_m(b))}{S_n(f)} df.
\end{align}
Note that the sum within each bin should be done with the same sampling rate as
the data, i.e. the same one would do in the case without using heterodyned
likelihood.

@@

In MCMC, the posterior distribution is approximated by a Markov chain that
converges to the target distribution. The Markov chain is constructed by
iteratively proposing a new point in the parameter space based on the current
location of the chain. The proposed point is accepted with a probability that is
usually set to be proportional to the ratio of the posterior density evaluated
at the proposed point and the current point. The chain can either accept the
proposal and move to the new location or reject the proposal and stay at the
current location. This process is repeated until the chain converges to the
target distribution. The samples generated by the chain are then used to
estimate the quantities of interest, such as the mean and credible intervals of
the source parameters. In practice, since we do not know the target distribution
ahead of time, the MCMC process is usually repeated until a certain criterion is
met, such as a Gelman-Rubin convergence statistic \cite{Gelman-rhat} lower than
a threshold or simply after a certain number of iterations.

Compared to direct sampling, MCMC algorithms only explore regions that are
highly probable, thus reducing the computational cost by not wasting resources in
regions where it is unlikely to generate the data. However, MCMC algorithms come
with their own set of issues. To illustrate what difficulties an MCMC algorithm
may face, we can examine one of the most vanilla MCMC algorithm, the
Metropolis-Hastings algorithm with a Gaussian kernel. Starting at some initial
point, one can draw a proposed point from a Gaussian transition kernel, defined
as
\begin{align}
    q(\mathbf{x},\mathbf{x_0})= \mathcal{N}(\mathbf{x}|\mathbf{x_0},\mathbf{C}),
\end{align}
where $\mathbf{x_0}$ is the current location of the chain, $\mathbf{x}$ is the
proposed location, and $\mathbf{C}$ is the covariance matrix of the Gaussian. In
the simplest case, we can pick $\mathbf{C}$ to be a diagonal matrix with a
constant value, which corresponds to an isotropic Gaussian center around the
current location with a constant variance. And the acceptance criteria is
defined as
\begin{align}
\alpha(\mathbf{x},\mathbf{x_0}) = \min\left(1,\frac{p(\mathbf{x})q(\mathbf{x_0},\mathbf{x})}{p(\mathbf{x_0})q(\mathbf{x},\mathbf{x_0})}\right).
\label{eq:Gaussian_acceptance}
\end{align}

We can see from eq. \ref{eq:Gaussian_acceptance} the acceptance rate is
proportion to the fraction of volume where the posterior density at the proposed
location is higher than the current location within the Gaussian transition
kernel. If we choose the variance of the transition kernel to be too large, this
fraction will be small hence the acceptance rate will be poor. On the other
hand, if one chooses the variance to be too small, nearby samples will be
correlated. In both cases, the efficiency in constructing the chain with a
target number of independent samples is suboptimal. There is often a tuning
process before we run the MCMC algorithm to find the optimal tuning parameters,
in this example the variance of the Gaussian, to ensure close to optimal
performance.

However, as we are dealing with higher dimensional problems, even the optimally
tuned Gaussian transition kernel does not guarantee good performance. In order
to have a reasonable acceptance rate, the variance of the Gaussian has to be
smaller in a higher dimensional space, which means the transition kernel is in
general making smaller and smaller steps as we increase the dimensionality.

Transition kernels that leverage gradient information of the target
distribution can help to address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluation at the current location to propose a new
point. For example, Metropolis-adjusted Langevin algorithm (MALA) \kw{cite}
place a unit Gaussian at the tip of the gradient vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I}),
\end{align}
where $\tau$ is the step size, which is a tuning parameter. Compared to a
Gaussian centers at the current location, the MALA transition kernel is more
likely to propose a point into the higher posterior density region because of
the gradient term, which helps to boost the acceptance rate.

While transition kernels that use gradient information can help to improve the
acceptance rate, computing the gradient of the posterior density function
introduce additional computational cost, which is not necessarily beneficial in
terms of sampling time. If one wants to compute the gradient information through
finite differencing, the additional computation cost goes as at least $\sim
\mathcal{O}(2n)$, where $n$ is the dimension of the problem. \texttt{Jax} allows
us to compute the gradient of the likelihood function with respect to the
parameters through automatic differentiation, which gives the gradient
information down to machine precision at around the same order of time compared
to evaluating the posterior itself. Having access to gradient information
through automatic differentiation is crucial to make the trade-off between using
gradient-based transition kernels and additional computation cost favorable.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples\kw{cite}, there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first-order derivatives here. Sampling
algorithms that use the information of higher order derivatives such as manifold-MALA
and Riemannian-HMC can in principle decorrelate local correlation in the target
distribution, however, they often have instability issues when they are used on
real-life applications, so they are not used often in practice.} For example,
target distribution that exhibits local correlation is hard to deal with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely 
be trapped in one mode and take an extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

% Long burn-in too
Moreover, before we can use the sampling chain to estimate the quantity we care about,
the sampler often needs to first find the most probable region in the target
space, which is a common process that is referred to as "burn-in" in the
literature. This means one would discard a certain amount of data generated from
the beginning and only use the later part to estimate the quantities of
interest. The burn-in phase of a gradient-based sampler is often as long as the
sampling phase, which means a good portion of computation is not necessarily
helpful in estimating the target quantities.

% Crux of normalizing flow
Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution and $f$ is an invertible parameterized transform that
connect the two distributions. For a detailed discussion of the algorithm, we
refer the readers to \kw{cite}.

% The flowMC algorithm
As mentioned in the previous subsection, one of the main issue gradient-based
sampler is it does not explore locally correlated features and multi-modality well. This
is exactly where the normalizing flow model can help. Once multiple independent
chains have generated samples, we can combine the samples and feed them to the
normalizing flow, which learns the global landscape of the target
distribution. Note that since the independent chains are generated with the same
number of samples, the normalization between posterior density represented by
different chains is in general different from the true target distribution. To
adjust for this inaccuracy, we use the normalizing flow model as a
\textit{proposal} distribution. Given a trained normalizing flow model, we can
generate the proposed jump in the target space by sampling from the latent
distribution $z \sim p_z(Z)$, then push it through the learned map given by the
normalizing flow model $x=f(z)$. the acceptance criteria is given by
\begin{align}
    \alpha(\mathbf{x},\mathbf{x_0}) = \min \left[ 1, \frac{\hat{\rho}(\mathbf{x_0})\rho_*(\mathbf{x})}{\hat{\rho}(\mathbf{x})\rho_*(\mathbf{x_0})}\right],
\end{align}
where $\hat{\rho}$ is the probability density estimated by the normalizing flow
model, $\rho_*$ is the probability density evaluated using the target function,
and $x_0$ is the current position. We can see the flow distribution is the
target distribution when the accepting probability is 1. When the normalizing
flow model has not converged to the target distribution, only a portion of the
proposed jumps will be accepted. This means an MCMC process using the normalizing
flow model as the proposal distribution can adjust the normalization across
different regions of the target parameter space by rejecting jumps into less
likely regions. The training and sampling are then repeated until certain
criteria are met. Note that every time we retrain the network, we are breaking the
Markov properties since we are changing the proposal distribution. To produce
final samples that can be used to estimate target quantities, one has to freeze
the normalizing flow model and not retrain during the final sampling phase.
To summarize the sampling process, the pseudocode is given by
